Practical Optional Types for Clojure

### 2015-02-26 formal-model.tex
section 4 better intro

We formalise our type system like \citet{TF10}
with differences highlighted in $\trdiff{\text{blue}}$.
The typing judgement 
$$
{\judgement   {\propenv}
              {\hastype {\e{}} {\t{}}}
  {\filterset {\thenprop {\prop{}}}
              {\elseprop {\prop{}}}}
  {\object{}}}
$$
says expression \e{} is of type \t{} in the 
proposition environment $\propenv{}$, with 
`then' proposition {\thenprop {\prop{}}}, `else' proposition {\elseprop {\prop{}}}
and object \object{}. We sometimes write 
{\judgementtwo{\propenv}{\hastype {\e{}} {\t{}}} if we are only interested in the type.

The propositions capture what can be assumed if an expression tests as logically true or false
respectively. The object is a possibly-empty symbolic representation of the expression.

@@

Types include the top type, untagged unions, functions, singleton types
and class instances. We abbreviate \EmptyUnion{} as \Bot{}, {\ValueNil} as \Nil{} and similarly
for \True and \False.
Function types contain \emph{latent} propositions and object, which, along with the return type,
may refer to the parameter identifier. They are latent because they are instantiated with the
actual object of the argument in applications before they are used in the proposition environment.

Propositions are standard---{\isprop {\t{}} {\path {\pathelem{}} {\x{}}}}
and {\notprop {\t{}} {\path {\pathelem{}} {\x{}}}}
say object {\path {\pathelem{}} {\x{}}} is and is not of type \t{}
respectively.
We include implications,
conjunctions, disjuctions, and the trivial and impossible propositions.
Proposition environments are sequences of propositions.

Figure~\ref{main:figure:typingrules} contains the standard typing rules.
The operational semantics are standard, except \nil{} is logically false:

@@

We extend our model with sequencing expressions and errors, where {\errorvalv{}}
is equivalent to Clojure's throw.

@@

Our main insight is as follows: 
if the left argument reduces to a value, then it is either logically true or false.
Occurrence typing gives exactly the propositions for both cases, so their disjunction
can be assumed for the right argument.
T-Do formalises this intuition.

\begin{mathpar}
    {\TDo}

    {\TError}
\end{mathpar}

The introduction of errors, specifically as an expression that is neither true nor false, 
makes our insight interesting.
For example, in checking \clj{(do (throw (Exception.)) (inc :a))}
by T-Error we assume \orprop{\botprop{}}{\botprop{}} in checking \clj{(inc :a)},
which is a contradiction by the logic system (L-False, figure~\ref{appendix:figure:proofsystem})
which allows us to assume \clj{(inc :a)} is unreachable.

Now we can check example~\ref{example:doexception}.
Intuitively, when evaluating
\clj{(inc x)} we know \clj{(assert (number? x))} has evaluated to a value,
so we assume \clj{x} is a number.
In detail, the example is checked first by T-Abs, which checks the body with the 
environment \propenv{} = {\isprop{\Union{\Nil}{\Number{}}}{\x{}}}.
Since the body is a \doliteral{},
T-Do extracts the propositions of the left argument, an if:
T-If says 
\thenprop{\prop{1}} = {\orprop{\botprop{}}{\notprop{\falsy{}}{\x{}}}}
by T-Nil and T-Subsume and
\elseprop{\prop{1}} = {\orprop{\botprop{}}{\botprop{}}} by T-Error.
Now {\orprop {\thenprop {\prop{1}}} {\elseprop {\prop{1}}}} = {\notprop{\falsy}{\x{}}}
and we check \clj{(inc x)} in the extended environment 
\ma{{\propenv{},{\notprop{\falsy}{\x{}}}}}.
By L-Update (figure~\ref{appendix:figure:proofsystem}) we know {\isprop{\Number}{\x{}}},
so T-Local checks \clj{x} as a \Number.

@@

@@

\subsection{\classconst{}}

The \classconst{} function is modeled as a constant which returns
the argument's class or \nil{}.

$$
\begin{array}{lrlr}
  \constantssyntax\\
  \pesyntax{}   &::=& \ldots \alt {\classpe{}}
                &\mbox{Path Elements}
\end{array}
$$

@@

The essential piece of occurrence typing's logical system is L-Update,
where we combine propositions that have similar objects to learn new facts.
(The metavariable \propisnotmeta{} ranges over \t{} and \nottype{\t{}} (without variables).)

@@

Now that we have our path elements \classpe{} and \keype{\k{}}
and the typing rules that introduce them, we give a detailed example how
each path element interacts with T-Update.

@@

From checking the body with T-Abs we know {\isprop {\Union{\Keyword}{\Nil}} {\x{}}}
The test has the object {\path{\classpe{}}{\x{}}} by T-App.
By T-If we can assume
{\notprop {\falsy{}} {\path{\classpe{}}{\x{}}}} in the then branch.
Since the object of the first proposition {\x{}} is a prefix of the second proposition's object
{\path{\classpe{}}{\x{}}}, 
by T-Update 
$$
\inpropenv {\propenv{}} {\isprop {\update {\Union{\Keyword}{\Nil}} {\nottype {\falsy{}}} {\classpe{}}} {\x{}}}.
$$
This hits the sixth \classpe{} case of \updateliteral{}, which returns
$$
{\update {\Union{\Keyword}{\Nil}} {\Object} {\emptypath{}}}
$$
deriving \inpropenv {\propenv{}} {\isprop {\Keyword} {\x{}}} 
by \restrictliteral{} (figure~\ref{main:figure:restrictremove}).

@@

T-Update combines these propositions
$$
\inpropenv {\propenv{}} {\isprop {\update {\HMapp{\mandatorysetsingleton{\makekw{kw}}{\Union{\Nil}{\Keyword}}}{\emptyabsent}} {\nottype {\falsy{}}} {\keype{\makekw{kw}}}} {\x{}}}.
$$
matching the \updateliteral{} case which pushes the \updateliteral{} inside the map
$$
{\HMapp{\mandatorysetsingleton{\makekw{kw}}{\update{\Union{\Nil}{\Keyword}}{\nottype {\falsy{}}}{\emptypath{}}}}{\emptyabsent}}
$$
concluding with
$$
\inpropenv {\propenv{}}
{\isprop {\HMapp{\mandatorysetsingleton{\makekw{kw}}{\Keyword}}{\emptyabsent}}   {\x{}}}.
$$

>>>>

Now that we have demonstrated the core features Typed Clojure
provides, we present a model that links them all together.
Our presentation will start with a review of
occurrence typing~\cite{TF10}.
Then for the rest of the section we incrementally add each
novel feature of Typed Clojure to the formalism,
interleaving presentation of typing rules, operational semantics
and
subtyping.

The first insight about occurrence typing is that
propositions can be used to capture type information about our programs,
in particular by referencing components or subcomponents of the runtime environment
and relating them to types.
\emph{Type Propositions} make assertions like ``variable \clj{x} is of type \clj{Number}'' or
``variable \clj{x} is not \clj{nil}''---
in our logical system we write these as
{\isprop{\Number}{\x{}}}
and {\notprop{\ValueNil}{\x{}}}. 
The other propositions are standard: the implication, conjunction,
disjunction, trivial (\topprop{}) and impossible (\botprop{}) propositions
(figure~\ref{main:figure:termsyntax}).

The particular part of the environment we reference in a positive or negative
type proposition is called the \emph{object}.
Every expression has an object that is either \emph{empty}, written \emptyobject{}, which says there is no
known way to relate this expression to the current environment, or a 
variable with some \emph{path}, written \path{\pathelem{}}{\x{}},
that exactly indicates how the value of this
expression can be derived from the current environment.
Type propositions can only reference non-empty objects.

The second insight is that we can replace the traditional 
representation of a
type environment of a map from variables to types
with a sequence of propositions, written \propenv{}. 
Instead of mapping \clj{x} to
the type \clj{Number}, we instead use the proposition {\isprop{\Number}{\x{}}}.

The third insight is that we
can use a logic system to learn new information about our programs.
The key component of the logic system L-Update uses type propositions
to derive new propositions.
$$
  {\LUpdate}
$$
For example, with L-Update we can use the knowledge of
\inpropenv{\propenv{}}{\isprop{\Union{\ValueNil{}}{\Number{}}}{\x{}}}
and 
\inpropenv{\propenv{}}{\notprop{\ValueNil{}}{\x{}}}
to derive \inpropenv{\propenv{}}{\isprop{\Number}{\x{}}}.
(The metavariable \propisnotmeta{} ranges over \t{} and \nottype{\t{}} (without variables).)
We cover \updateliteral{} in section~\ref{sec:updateformalsection}.

The fourth insight is a consequence of
programmers in dynamic languages
using implicit type-based reasoning based on the result of
conditional tests.
For instance, example~\ref{example:conditionalflow}
only utilises \clj{x} once
the programmer is convinced it is safe to do so based whether
\clj{(number? x)}
is 
a true or false value. The takeaway is that every expression has
two relevant pieces of information to the logic system: a `then' proposition
for when it reduces to a true value, and an `else' proposition
when it reduces to a false value---for \clj{(number? x)}
they are {\isprop{\Number}{\x{}}} and {\notprop{\Number}{\x{}}}
respectively.
%\ref{main:figure:typingrules}

We formalise our type system following~\citet{TF10}
(with differences highlighted in $\trdiff{\text{blue}}$).
The typing judgement 
$$
{\judgement   {\propenv}
              {\hastype {\e{}} {\t{}}}
  {\filterset {\thenprop {\prop{}}}
              {\elseprop {\prop{}}}}
  {\object{}}}
$$
says expression \e{} is of type \t{} in the 
proposition environment $\propenv{}$, with 
`then' proposition {\thenprop {\prop{}}}, `else' proposition {\elseprop {\prop{}}}
and object \object{}. We sometimes write 
{\judgementtwo{\propenv}{\hastype {\e{}} {\t{}}} if we are only interested in the type.

@@

Types include the top type, \emph{untagged} unions, functions, singleton types
and class instances. We abbreviate \EmptyUnion{} as \Bot{}, {\ValueNil} as \Nil{}, 
{\ValueTrue} as \True and {\ValueFalse} as {\False}.
Function types contain \emph{latent}(terminology from~\cite{Lucassen88polymorphiceffect}) propositions and object, which, along with the return type,
may refer to the parameter identifier. They are latent because they are instantiated with the
actual object of the argument in applications before they are used in the proposition environment.

Figure~\ref{main:figure:typingrules} contains the standard typing rules.
We can now demonstrate how to check example~\ref{example:conditionalflow}.
$$
\judgement{\isprop{\clj{(U nil Num)}}{\x{}}}{\hastype{\clj{(number? x)}}{\clj{Boolean}}}{\filterset{\isprop{\clj{Num}}{\x{}}}{\notprop{\clj{Num}}{\x{}}}}{\emptyobject{}}
$$
$$
\judgement{\isprop{\clj{Num}}{\x{}}}{\hastype{\clj{(inc x)}}{\clj{Num}}}{\filterset{\topprop{}}{\topprop{}}}{\emptyobject{}}
$$
$$
\judgement{\isprop{\clj{Num}}{\x{}}}{\hastype{\clj{x}}{\clj{Num}}}{\filterset{\topprop{}}{\topprop{}}}{\x{}}
$$
$$
\judgement{\isprop{\clj{Num}}{\x{}}}{\hastype{\clj{0}}{\clj{Num}}}{\filterset{\topprop{}}{\topprop{}}}{\emptyobject{}}
$$
By T-Abs

The operational semantics are standard, except \nil{} is logically false:

@@

We extend our model with sequencing expressions and errors, where {\errorvalv{}}
models Clojure's throw special form.

@@

Our main insight is as follows: 
if the left argument reduces to a value, then it is either logically true or false.
If we learn the same information in both cases then we can use it as an assumption to check the right argument.
T-Do formalises this intuition.

\begin{mathpar}
    {\TDo}  
\end{mathpar}

The introduction of errors, specifically as an expression that is neither true nor false, 
makes our insight interesting.

\begin{mathpar}
    {\TError}
\end{mathpar}

Now we can check example~\ref{example:doexception}.
We check the test with T-App.
$$
\judgement{\isprop{\clj{(U nil Num)}}{\x{}}}{\hastype{\clj{(number? x)}}{\clj{B}}}{\filterset{\isprop{\clj{Num}}{\x{}}}{\notprop{\clj{Num}}{\x{}}}}{\emptyobject{}}
$$
By T-Nil and subsumption we derive the same information in both propositions.
$$
\judgement{\isprop{\clj{Num}}{\x{}}}{\hastype{\clj{nil}}{\clj{nil}}}{\filterset{\isprop{\clj{Num}}{\x{}}}{\notprop{\clj{Num}}{\x{}}}}{\emptyobject{}}
$$
By T-Error we can assume anything about the return of an error.
$$
\judgement{\notprop{\clj{Num}}{\x{}}}{\hastype{\errorvalv{}}{\Bot}}{\filterset{\botprop{}}{\botprop{}}}{\emptyobject{}}
$$
Using the above as premises to T-If we learn {\orprop{\isprop{\clj{Num}}{\x{}}}{\botprop{}}}
in both propositions.
$$
\judgement{\isprop{\clj{(U nil Num)}}{\x{}}}{\hastype{\clj{(if (number? x) nil}\ {\errorvalv{}{\clj{)}}}}{\clj{B}}}
          {\filterset{\orprop{\isprop{\clj{Num}}{\x{}}}{\botprop{}}}{\orprop{\notprop{\clj{Num}}{\x{}}}{\botprop{}}}}{\emptyobject{}}
$$
By T-Local we know \clj{x} is a \clj{Num}.
$$
\judgement{\orprop{\isprop{\clj{Num}}{\x{}}}{\botprop{}}}{\hastype{\clj{x}}{\clj{Num}}}{\filterset{\topprop{}}{\topprop{}}}{\x{}}
$$
\clj{inc} requires a \clj{Num} argument which we just proved.
$$
\judgement{\orprop{\isprop{\clj{Num}}{\x{}}}{\botprop{}}}{\hastype{\clj{(inc x)}}{\clj{Num}}}{\filterset{\topprop{}}{\topprop{}}}{\emptyobject{}}
$$

@@

\subsection{Paths}

@@

The \classconst{} function is a constant which returns
the argument's class or \nil{}.

$$
\begin{array}{lrlr}
  \pesyntax{}   &::=& \ldots \alt {\classpe{}}
                &\mbox{Path Elements}
\end{array}
$$

@@

The essential piece of occurrence typing's logical system is L-Update,
where we combine propositions that have similar objects to learn new facts.

@@

Now that we have our path elements \classpe{} and \keype{\k{}}
and the typing rules that introduce them, we give a detailed example how
each path element interacts with L-Update.

@@

From checking the body with T-Abs we know {\isprop {\Union{\Keyword}{\Nil}} {\x{}}}
The test has the object {\path{\classpe{}}{\x{}}} by T-App.
By T-If we can assume
{\notprop {\falsy{}} {\path{\classpe{}}{\x{}}}} in the then branch.
Since the object of the first proposition {\x{}} is a prefix of the second proposition's object
{\path{\classpe{}}{\x{}}}, 
by L-Update 
$$
\inpropenv {\propenv{}} {\isprop {\update {\Union{\Keyword}{\Nil}} {\nottype {\falsy{}}} {\classpe{}}} {\x{}}}.
$$
This hits the sixth \classpe{} case of \updateliteral{}, which returns
$$
{\update {\Union{\Keyword}{\Nil}} {\Object} {\emptypath{}}}
$$
deriving \inpropenv {\propenv{}} {\isprop {\Keyword} {\x{}}} 
by \restrictliteral{} (figure~\ref{main:figure:restrictremove}).

@@

L-Update combines these propositions
$$
\inpropenv {\propenv{}} {\isprop {\update {\HMapp{\mandatorysetsingleton{\makekw{kw}}{\Union{\Nil}{\Keyword}}}{\emptyabsent}} {\nottype {\falsy{}}} {\keype{\makekw{kw}}}} {\x{}}}.
$$
matching the \updateliteral{} case which pushes the \updateliteral{} inside the map
$$
{\HMapp{\mandatorysetsingleton{\makekw{kw}}{\update{\Union{\Nil}{\Keyword}}{\nottype {\falsy{}}}{\emptypath{}}}}{\emptyabsent}}
$$
concluding with
$$
\inpropenv {\propenv{}}
{\isprop {\HMapp{\mandatorysetsingleton{\makekw{kw}}{\Keyword}}{\emptyabsent}}   {\x{}}}.
$$

### 2015-02-26 overview.tex
section 4 better intro

We introduce Clojure's syntax.
Parenthesis indicate \emph{applications}, square brackets
are syntax for literal \emph{vectors} and curly braces delimit literal
\emph{hash-maps}.
\clj{nil} is exactly Java's \java{null} and is one of two logically
false values (the other is \clj{false}).
\emph{Symbols} begin with an alphabetic character.
A symbol prefixed with a colon like \clj{:a} is an interned \emph{keyword}.
The \clj{get} function takes a hash-map and a key returns the value
associated with the key in the map, or returns \clj{nil} if there is none.
Keywords are functions that take a map and applies \clj{get} to the map
and itself.
\clj{inc} and \clj{dec} increment and decrement numbers respectively.
\clj{assoc} and \clj{dissoc} associate and dissociate entries
from immutable maps.
Everything is an instance of some class
except for \clj{nil}. Predicates like \clj{number?} are equivalent to
a Java \java{instanceof} test, in this case on \clj{java.lang.Number}.

@@

Clojure uses \emph{namespaces}, declared with the
\clj{ns} form, to manage top-level \emph{vars} and
class imports.
Example~\ref{example:conditionalflow}
is a regular Clojure file compiled with
the Clojure compiler except it declares a runtime dependency on
\clj{clojure.core.typed}, Typed Clojure's core namespace.
Programmers import typed versions of certain constructs as needed, like
a \clj{fn} variant that supports annotations,
and use the provided \clj{check-ns} function to type check the current namespace
when convenient.
We omit \clj{ns} forms for the rest of the examples.
\footnote{See <this> github repo for full working examples}

@@

In a static type, a quoted keyword \clj{':if} is a singleton type containing just the keyword.
Quoted maps \clj{'{:op ':if}} are HMap types---the keys are not quoted because only keyword keys
are allowed. The type constructor \clj{U} indicates an untagged union type, which can take
any number of types.

>>>>

We introduce Clojure's syntax.
Parenthesis indicate \emph{applications}, square brackets
are syntax for literal \emph{vectors} and curly braces delimit literal
\emph{hash-maps}.
\clj{nil} is exactly Java's \java{null} and is one of two logically
false values (the other is \clj{false}).
\emph{Symbols} begin with an alphabetic character.
A symbol prefixed with a colon like \clj{:a} is an interned \emph{keyword}.
The \clj{get} function takes a hash-map and a key returns the value
associated with the key in the map, or returns \clj{nil} if there is none.
Keywords are functions that take a map and applies \clj{get} to the map
and itself.
\clj{inc} and \clj{dec} increment and decrement numbers respectively.
\clj{assoc} and \clj{dissoc} associate and dissociate entries
from immutable maps.
The \clj{=} function returns true if all its arguments are structurally equal---in this
paper we treat it as pointer equality.
Everything is an instance of some class
except for \clj{nil}. Predicates like \clj{number?} are equivalent to
a Java \java{instanceof} test, in this case on \clj{java.lang.Number}.

@@

Clojure uses \emph{namespaces}, declared with the
\clj{ns} form, to manage top-level \emph{vars} and
class imports.
Example~\ref{example:conditionalflow}
is a regular Clojure file compiled with
the Clojure compiler using \clj{:require} to declare a
runtime dependency on Typed Clojure's core namespace
\clj{clojure.core.typed}.
Programmers import typed versions of certain constructs as needed, like
a \clj{fn} variant that supports annotations,
and use the provided \clj{check-ns} function to type check the current namespace
when convenient.
We omit \clj{ns} forms for the rest of the examples.
\footnote{See <this> github repo for full working examples}

@@

In a static type, a quoted keyword \clj{':if} is a singleton type containing just the keyword.
Quoted maps like \clj{'{:op ':const, :val Num}} are HMap types representing
maps with known keyword entries with values of the associated type
The keys are not quoted because only keyword keys
are allowed. The type constructor \clj{U} indicates an untagged union type, which can take
any number of types.

The \clj{defalias} macro defines a type abbreviation which can be recursively defined.
